---
title: "Meta-analysis of prediction model performance"
author:
  - name: Thomas Debray
    orcid: 0000-0002-1790-2719
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Meta-analysis of prediction model performance}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
bibliography: "https://api.citedrive.com/bib/0d25b38b-db8f-43c4-b934-f4e2f3bd655a/references.bib?x=eyJpZCI6ICIwZDI1YjM4Yi1kYjhmLTQzYzQtYjkzNC1mNGUyZjNiZDY1NWEiLCAidXNlciI6ICIyNTA2IiwgInNpZ25hdHVyZSI6ICI0MGFkYjZhMzYyYWE5Y2U0MjQ2NWE2ZTQzNjlhMWY3NTk5MzhhNzUxZDNjYWIxNDlmYjM4NDgwOTYzMzY5YzFlIn0=/bibliography.bib"
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```


# Case Study
The [EuroSCORE II](http://www.euroscore.org/calc.html) is a commonly used scoring rule for estimating the risk of  in-hospital mortality in  patients undergoing major cardiac surgery. It was developed using individual participant data from  16,828 adult patients from 154 hospitals across 43 countries. Predictors include patient characteristics (e.g. age, gender), cardiac related factors (e.g. recent MI) and surgery related factors (e.g. Surgery on thoracic aorta). In 2014, a systematic review of published evidence on the performance value of the EuroSCORE II was undertaken by @guida_performance_2014. Twenty-two validations, including 145,592 patients from 21 external validation articles (one study included two validations) and a split-sample validation contained within original development article were included in the review; 23 validation studies in total.

In this vignette, we illustrate the R package **metamisc** to summarize the results from aformentioned 23 validation studies. In particular, we will derive summary estimates for the discrimination and calibration performance of EuroSCORE II, evaluate the presence of between-study heterogeneity, and assess to which extent study characteristics influence its performance. 

The methodology of this package is explained and illustrated by @debray_guide_2017. 

```{r setup}
library(metamisc)
```

# References
